---
title: "Case study 3 : IGF case study"
author: "Sara Taheri"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(rstan)
library(ggplot2)
library(ggplot2)
library(tidyr)
library(tibble)
library(smfsb)
library(dplyr)
```

```{r}
D = 300
startSeed = 1
K = 10
ndp <- c(30, 60, 100, 140, 200) #number of data points
```


# Create observational and interventional data

```{r}

#####################################################################################################
######################################### SDE #######################################################
#####################################################################################################
# PRE <- as.matrix(read.csv('growth_factor_sheets/growth_factor/Values-Pre.csv', header = TRUE))
# POST <- as.matrix(read.csv('growth_factor_sheets/growth_factor/Values-Post.csv', header = TRUE))
# 
# gf_sde <- function(states, rates, interventions = NULL){
#   sde <- list()
#   
#   sde$Pre <- PRE
#   sde$Post <- POST
#   
#   innerIntervention <- interventions
#   
#   sde$h <- function(states, t, parameters=rates, interventions = innerIntervention){
#     # update the initial states
#     if(!is.null(interventions)) {
#       for(int in names(interventions)){
#         states[[int]] <- interventions[[int]]
#       }
#     }
#     with(as.list(c(states, parameters, interventions)), {
#       if(!is.null(interventions)) {
#         for(int in names(interventions)){
#           sde$Pre[,int] <- 0
#           sde$Post[,int] <- 0
#         }
#       }
#       out <- c(
#         SOS_activation_by_EGFR * SOS_inactive * EGFR,
#         SOS_activation_by_IGFR * SOS_inactive * IGFR,
#         SOS_deactivation * SOS_active,
#         Ras_activation_by_SOS * Ras_inactive * SOS_active,
#         Ras_deactivation * Ras_active,
#         PI3K_activation_by_EGFR * PI3K_inactive * EGFR,
#         PI3K_activation_by_IGFR * PI3K_inactive * IGFR,
#         PI3K_activation_by_Ras * PI3K_inactive * Ras_active,
#         PI3K_deactivation * PI3K_active,
#         AKT_activation_by_PI3K * AKT_inactive * PI3K_active,
#         AKT_deactivation * AKT_active,
#         Raf_activation_by_Ras * Raf_inactive * Ras_active,
#         Raf_deactivation_by_phosphotase * Raf_active,
#         Raf_deactivation_by_AKT * AKT_active * Raf_active,
#         Mek_activation_by_Raf * Mek_inactive * Raf_active,
#         Mek_deactivation * Mek_active,
#         Erk_activation_by_Mek * Erk_inactive * Mek_active,
#         Erk_deactivation * Erk_active
#       )
#       names(out) <- c("SOS_inactive_to_SOS_active","SOS_inactive_to_SOS_active","SOS_active_to_SOS_inactive",
#                        "Ras_inactive_to_Ras_active", "Ras_active_to_Ras_inactive",
#                       "PI3K_inactive_to_PI3K_active","PI3K_inactive_to_PI3K_active","PI3K_inactive_to_PI3K_active","PI3K_active_to_PI3K_inactive",
#                       "AKT_inactive_to_AKT_active","AKT_active_to_AKT_inactive",
#                       "Raf_inactive_to_Raf_active", "Raf_active_to_Raf_inactive", "Raf_active_to_Raf_inactive",
#                       "Mek_inactive_to_Mek_active", "Mek_active_to_Mek_inactive",
#                       "Erk_inactive_to_Erk_active", "Erk_active_to_Erk_inactive")
#       if(!is.null(interventions)) {
#         for(int in names(interventions)){
#           out[which(grepl(int, names(out), ignore.case = TRUE) == TRUE)] <- 0
#         }
#       }
#       out <- unname(out)
#       
#       return(out)
#     })
#   }
#   transition_function <- StepGillespie(sde)
#   return(transition_function)
# }
# 
# sde_sim <- function(transition_function, initial_states, times, interventions = NULL) {
#   if(!is.null(interventions)) {
#     for(int in names(interventions)){
#       initial_states[[int]] <- interventions[[int]]
#     }
#   }
#   initial_states <- structure(as.numeric(initial_states), names = names(initial_states))
#   t_delta <- times[2] - times[1]
#   out <- as_tibble(
#     smfsb::simTs(initial_states, times[1], times[length(times)], t_delta, transition_function)
#   )
#   out$time <- times[0:(length(times)-1)]
#   out <- out[, c('time', setdiff(names(out), 'time'))]
#   return(out)
# }
# 
# # create observational and interventional data from SDE when number of phosphorylated SOS = 70
# 
# rates <- list(
#   SOS_activation_by_EGFR=.01,
#   SOS_activation_by_IGFR=.01,
#   SOS_deactivation=.5,
#   Ras_activation_by_SOS=.01,
#   Ras_deactivation=.5,
#   PI3K_activation_by_EGFR=.01,
#   PI3K_activation_by_IGFR=.01,
#   PI3K_activation_by_Ras=.01,
#   PI3K_deactivation=.5,
#   AKT_activation_by_PI3K=.01,
#   AKT_deactivation=.5,
#   Raf_activation_by_Ras=.01,
#   Raf_deactivation_by_AKT=.01,
#   Raf_deactivation_by_phosphotase=.3,
#   Mek_activation_by_Raf=.05,
#   Mek_deactivation=.5,
#   Erk_activation_by_Mek=.05,
#   Erk_deactivation=.5
# )
# 
# observational_data_list <- list()
# intervention_data_list <- list()
# 
# for (seed in startSeed:(startSeed+K-1)) {
#   set.seed(seed)
#   observational_data <- data.frame("time" = 0, "EGFR" = 0, "IGFR" = 0, "SOS_inactive" = 0, "SOS_active" = 0,
#                                    "Ras_inactive" = 0, "Ras_active" = 0,"PI3K_inactive" = 0, "PI3K_active" = 0,
#                                    "AKT_inactive" = 0, "AKT_active" = 0,"Raf_inactive" = 0, "Raf_active" = 0,
#                                    "Mek_inactive" = 0,"Mek_active" = 0, "Erk_inactive" = 0, "Erk_active" = 0)
#   intervention_data <- observational_data
#   
#   EGFR_r <- rnorm(D,37,3)
#   IGFR_r <- rnorm(D,5,1)
#   for (i in 1:D) {
#     initial_states <-  list(
#       EGFR=EGFR_r[i],
#       IGFR=IGFR_r[i],
#       SOS_inactive=100,
#       SOS_active=0,
#       Ras_inactive=100,
#       Ras_active=0,
#       PI3K_inactive=100,
#       PI3K_active=0,
#       AKT_inactive=100,
#       AKT_active=0,
#       Raf_inactive=100,
#       Raf_active=0,
#       Mek_inactive=100,
#       Mek_active=0,
#       Erk_inactive=100,
#       Erk_active=0
#     )
#     set.seed(i)
#     times <- seq(0, 1, by = .1)
#     faster_rates <- lapply(rates, `*`, 20)
#     
#     stoc_transition_func <- gf_sde(initial_states, faster_rates,interventions = NULL)
#     sde_out <- sde_sim(stoc_transition_func, initial_states, times,interventions = NULL)
#     observational_data <- rbind(observational_data,sde_out[nrow(sde_out),])
#     
#     stoc_transition_func <- gf_sde(initial_states, faster_rates,interventions = list(SOS_inactive=30, SOS_active=70))
#     sde_out <- sde_sim(stoc_transition_func, initial_states, times,interventions = list(SOS_inactive=30, SOS_active=70))
#     intervention_data <- rbind(intervention_data,sde_out[nrow(sde_out),])
#   }
#   observational_data <- observational_data[,c("EGFR","IGFR","SOS_active","Ras_active","PI3K_active","AKT_active","Raf_active","Mek_active","Erk_active")]
#   colnames(observational_data) <- c("EGFR","IGFR","SOS","Ras","PI3K","AKT","Raf","Mek","Erk")
#   observational_data <- observational_data[-1,]
#   rownames(observational_data) <- seq(1:D)
#   
#   intervention_data <- intervention_data[,c("EGFR","IGFR","SOS_active","Ras_active","PI3K_active","AKT_active","Raf_active","Mek_active","Erk_active")]
#   colnames(intervention_data) <- c("EGFR","IGFR","SOS","Ras","PI3K","AKT","Raf","Mek","Erk")
#   intervention_data <- intervention_data[-1,]
#   rownames(intervention_data) <- seq(1:D)
#   
#   observational_data_list[[seed]] <- observational_data
#   intervention_data_list[[seed]] <- intervention_data
# }
# 
# saveRDS(observational_data_list,"data/Case3/observationalD_igfSOS70.RData") #save observational data
# saveRDS(intervention_data_list,"data/Case3/interventionD_igf_SOS70.RData") #save intervention data

```

Once we create the data, we will save them in data folder and from now on, we only read the data from the data folder.

```{r}
observational_data_list <- readRDS("data/Case3/observationalD_igfSOS70.RData")
intervention_data_list <- readRDS("data/Case3/interventionD_igf_SOS70.RData")
```

# IGF case study Stan model where IGF, EGF and PI3K are hidden

```{r}
model_str_IGF <- "
    data {
        int D;
        vector[D] SOS_train;
        vector[D] AKT_train;
        vector[D] Raf_train;
        vector[D] Mek_train;
        vector[D] Erk_train;
        vector[D] Ras_train;
    }
  parameters {
       real<lower=0> mu_EGF; 
       real<lower=0> sigma_EGF;
       real<lower=0> mu_IGF; 
       real<lower=0> sigma_IGF;
       real beta0_EGFIGFToSOS; 
       real<lower=0> beta_EGFToSOS;
       real<lower=0> beta_IGFToSOS;
       real beta0_SOSToRas;
       real<lower=0> beta_SOSToRas;
       real beta0_AKT;
       real<lower=0> beta_EGFToAKT;
       real<lower=0> beta_IGFToAKT;
       real<lower=0> beta_RasToAKT;
       real beta0_RasAKTToRaf;
       real<lower=0> beta_RasToRaf;
       real<upper=0> beta_AKTToRaf;
       real beta0_RafToMek;
       real<lower=0> beta_RafToMek;
       real beta0_MekT0Erk;
       real<lower=0> beta_MekT0Erk;
       vector[D] EGF_train;
       vector[D] IGF_train;
    }
    transformed parameters {
      vector[D] SOS_loc;
      vector[D] Ras_loc;
      vector[D] AKT_loc;
      vector[D] Raf_loc;
      vector[D] Mek_loc;
      vector[D] Erk_loc;
      for (i in 1:D){
        SOS_loc[i] = 100 / (1 + exp(-beta0_EGFIGFToSOS - EGF_train[i] * beta_EGFToSOS - IGF_train[i] * beta_IGFToSOS));
        Ras_loc[i] = 100 / (1 + exp(-beta0_SOSToRas -SOS_train[i] * beta_SOSToRas));
        AKT_loc[i] = 100 / (1 + exp(-beta0_AKT - EGF_train[i] * beta_EGFToAKT - IGF_train[i] * beta_IGFToAKT - Ras_train[i] * beta_RasToAKT));
        Raf_loc[i] = 100 / (1 + exp(-beta0_RasAKTToRaf - Ras_train[i] * beta_RasToRaf - AKT_train[i] * beta_AKTToRaf));
        Mek_loc[i] = 100 / (1 + exp(-beta0_RafToMek - Raf_train[i] * beta_RafToMek));
        Erk_loc[i] = 100 / (1 + exp(-beta0_MekT0Erk - Mek_train[i] * beta_MekT0Erk));
      }
    }
    model {
        beta0_EGFIGFToSOS  ~ normal(0, 10); 
        beta_EGFToSOS  ~ normal(0, 10);
        beta_IGFToSOS  ~ normal(0, 10);
        mu_EGF ~ normal(37, 3);
        sigma_EGF ~ normal(1,1);
        mu_IGF ~ normal(5, 1);
        sigma_IGF ~ normal(1,1);
        beta0_SOSToRas ~ normal(0,10);
        beta_SOSToRas ~ normal(0, 10);
        beta0_AKT ~ normal(0,10);
        beta_EGFToAKT ~ normal(0,10);
        beta_IGFToAKT ~ normal(0,10);
        beta_RasToAKT ~ normal(0,10);
        beta0_RasAKTToRaf ~ normal(0,10);
        beta_RasToRaf ~ normal(0,10);
        beta_AKTToRaf ~ normal(0,10);
        beta0_RafToMek ~ normal(0,10);
        beta_RafToMek ~ normal(0,10);
        beta0_MekT0Erk ~ normal(0,10);
        beta_MekT0Erk ~ normal(0,10);
        EGF_train ~ normal(mu_EGF, sigma_EGF);
        IGF_train ~ normal(mu_IGF, sigma_IGF);
        SOS_train ~ normal(SOS_loc, 5.4);      // likelihood
        Ras_train ~ normal(Ras_loc, 5.3); //ras_loc = 47
        AKT_train ~ normal(AKT_loc,5.1);
        Raf_train ~ normal(Raf_loc,5.3);
        Mek_train ~ normal(Mek_loc,4.7);
        Erk_train ~ normal(Erk_loc,3.2);
    }
"
```

```{r, message=FALSE, warning=FALSE}
mod <- rstan::stan_model(model_code = model_str_IGF)
```

# Mutilated model

```{r}
num_samples = 500
mutilated_model <- function(mu_IGF, sigma_IGF, mu_EGF, sigma_EGF
                            ,beta0_EGFIGFToSOS, beta_EGFToSOS, beta_IGFToSOS
                            ,beta0_SOSToRas, beta_SOSToRas
                            ,beta0_AKT,beta_EGFToAKT, beta_IGFToAKT, beta_RasToAKT
                            ,beta0_RasAKTToRaf, beta_RasToRaf, beta_AKTToRaf
                            ,beta0_RafToMek, beta_RafToMek
                            ,beta0_MekT0Erk, beta_MekT0Erk
                            ,sos
                            ) {
  EGF = rnorm(num_samples, mu_EGF, sigma_EGF)
  IGF = rnorm(num_samples, mu_IGF, sigma_IGF)
  SOS = rep(sos, num_samples)
  Ras = 100 / (1 + exp(-beta0_SOSToRas - SOS * beta_SOSToRas)) + rnorm(num_samples,0,5.5)
  AKT = 100 / (1 + exp(-beta0_AKT - EGF * beta_EGFToAKT - IGF * beta_IGFToAKT - Ras * beta_RasToAKT)) + rnorm(n = num_samples,0,5.2)
  Raf = 100 / (1 + exp(-beta0_RasAKTToRaf - Ras * beta_RasToRaf - AKT * beta_AKTToRaf)) + rnorm(n = num_samples,0,5.4)
  Mek = 100 / (1 + exp(-beta0_RafToMek - Raf * beta_RafToMek)) + rnorm(n = num_samples,0,4.9)
  Erk = 100 / (1 + exp(-beta0_MekT0Erk - Mek * beta_MekT0Erk)) + rnorm(n = num_samples,0,3.2)
  return(Erk)
}
```


```{r, echo = FALSE}
estimated_y_hmc_list <- list()
hmc_fit_list <- list()
means_hmc <- c()
means_hmc_list <- list()
start_time <- Sys.time()
for (num_data_points in ndp) {
  for (data_count in 1:K) {
    data_list <- list(D=num_data_points
                  ,SOS_train = observational_data_list[[data_count]]$SOS[1:num_data_points]
                  ,AKT_train = observational_data_list[[data_count]]$AKT[1:num_data_points]
                  ,Raf_train = observational_data_list[[data_count]]$Raf[1:num_data_points]
                  ,Mek_train = observational_data_list[[data_count]]$Mek[1:num_data_points]
                  ,Erk_train = observational_data_list[[data_count]]$Erk[1:num_data_points]
                  ,Ras_train = observational_data_list[[data_count]]$Ras[1:num_data_points]
                  )
    
    #If you get this error (Cluster setup failed. 2 of 2 workers failed to connect.) run these lines:
    if (Sys.getenv("RSTUDIO") == "1" && !nzchar(Sys.getenv("RSTUDIO_TERM")) &&
        Sys.info()["sysname"] == "Darwin") {
      parallel:::setDefaultClusterOptions(setup_strategy = "sequential")
    }
    hmc_fit_list[[data_count]] <- rstan::sampling(mod, data=data_list, chains = 2, iter = 4000, warmup = 2000, seed = 1, control = list(max_treedepth = 15, adapt_delta = 0.99))
    samples_hmc <- rstan::extract(hmc_fit_list[[data_count]], c("mu_EGF", "sigma_EGF"
                                     ,"mu_IGF", "sigma_IGF"
                                     ,"beta0_EGFIGFToSOS","beta_EGFToSOS", "beta_IGFToSOS"
                                     ,"beta0_SOSToRas","beta_SOSToRas"
                                     ,"beta0_AKT", "beta_EGFToAKT", "beta_IGFToAKT", "beta_RasToAKT"
                                     ,"beta0_RasAKTToRaf", "beta_RasToRaf", "beta_AKTToRaf"
                                     ,"beta0_RafToMek", "beta_RafToMek"
                                     ,"beta0_MekT0Erk", "beta_MekT0Erk"
                                     ))
    estimated_y_hmc_list[[data_count]] <-  mutilated_model(mu_IGF = mean(samples_hmc$mu_IGF),
                                                                          sigma_IGF = mean(samples_hmc$sigma_IGF),
                                                                          mu_EGF = mean(samples_hmc$mu_EGF),
                                                                          sigma_EGF = mean(samples_hmc$sigma_EGF),
                                                                          beta0_EGFIGFToSOS = mean(samples_hmc$beta0_EGFIGFToSOS),
                                                                          beta_EGFToSOS = mean(samples_hmc$beta_EGFToSOS),
                                                                          beta_IGFToSOS = mean(samples_hmc$beta_IGFToSOS),
                                                                          beta0_SOSToRas = mean(samples_hmc$beta0_SOSToRas),
                                                                          beta_SOSToRas = mean(samples_hmc$beta_SOSToRas),
                                                                          beta0_AKT = mean(samples_hmc$beta0_AKT),
                                                                          beta_EGFToAKT = mean(samples_hmc$beta_EGFToAKT),
                                                                          beta_IGFToAKT = mean(samples_hmc$beta_IGFToAKT),
                                                                          beta_RasToAKT = mean(samples_hmc$beta_RasToAKT),
                                                                          beta0_RasAKTToRaf = mean(samples_hmc$beta0_RasAKTToRaf),
                                                                          beta_RasToRaf = mean(samples_hmc$beta_RasToRaf),
                                                                          beta_AKTToRaf = mean(samples_hmc$beta_AKTToRaf),
                                                                          beta0_RafToMek = mean(samples_hmc$beta0_RafToMek),
                                                                          beta_RafToMek = mean(samples_hmc$beta_RafToMek),
                                                                          beta0_MekT0Erk = mean(samples_hmc$beta0_MekT0Erk),
                                                                          beta_MekT0Erk = mean(samples_hmc$beta_MekT0Erk),
                                                                          sos = 70
    )
    means_hmc <- c(means_hmc,abs(mean(estimated_y_hmc_list[[data_count]]) - mean(intervention_data_list[[data_count]]$Erk)))
  #   
   }
  means_hmc_list[[paste0("num_data_points_",num_data_points)]] <- means_hmc
  means_hmc <- c()
}
end_time <- Sys.time()
end_time - start_time
```


# Misspecified IGF case study:  where EGF and PI3K are hidden and IGF is missed from the model

```{r}
model_str_IGF_misspecify <- "
    data {
        int D;
        vector[D] SOS_train;
        vector[D] AKT_train;
        vector[D] Raf_train;
        vector[D] Mek_train;
        vector[D] Erk_train;
        vector[D] Ras_train;
    }
  parameters {
       real<lower=0> mu_EGF; 
       real<lower=0> sigma_EGF;
       real beta0_EGFToSOS; 
       real<lower=0> beta_EGFToSOS;
       real beta0_SOSToRas;
       real<lower=0> beta_SOSToRas;
       real beta0_AKT;
       real<lower=0> beta_EGFToAKT;
       real<lower=0> beta_RasToAKT;
       real beta0_RasAKTToRaf;
       real<lower=0> beta_RasToRaf;
       real<upper=0> beta_AKTToRaf;
       real beta0_RafToMek;
       real<lower=0> beta_RafToMek;
       real beta0_MekT0Erk;
       real<lower=0> beta_MekT0Erk;
       vector[D] EGF_train;
    }
    transformed parameters {
      vector[D] SOS_loc;
      vector[D] Ras_loc;
      vector[D] AKT_loc;
      vector[D] Raf_loc;
      vector[D] Mek_loc;
      vector[D] Erk_loc;
      for (i in 1:D){
        SOS_loc[i] = 100 / (1 + exp(-beta0_EGFToSOS - EGF_train[i] * beta_EGFToSOS));
        Ras_loc[i] = 100 / (1 + exp(-beta0_SOSToRas -SOS_train[i] * beta_SOSToRas));
        AKT_loc[i] = 100 / (1 + exp(-beta0_AKT - EGF_train[i] * beta_EGFToAKT - Ras_train[i] * beta_RasToAKT));
        Raf_loc[i] = 100 / (1 + exp(-beta0_RasAKTToRaf - Ras_train[i] * beta_RasToRaf - AKT_train[i] * beta_AKTToRaf));
        Mek_loc[i] = 100 / (1 + exp(-beta0_RafToMek - Raf_train[i] * beta_RafToMek));
        Erk_loc[i] = 100 / (1 + exp(-beta0_MekT0Erk - Mek_train[i] * beta_MekT0Erk));
      }
    }
    model {
        beta0_EGFToSOS  ~ normal(0, 10); 
        beta_EGFToSOS  ~ normal(0, 10);
        mu_EGF ~ normal(37, 3);
        sigma_EGF ~ normal(1,1);
        beta0_SOSToRas ~ normal(0,10);
        beta_SOSToRas ~ normal(0, 10);
        beta0_AKT ~ normal(0,10);
        beta_EGFToAKT ~ normal(0,10);
        beta_RasToAKT ~ normal(0,10);
        beta0_RasAKTToRaf ~ normal(0,10);
        beta_RasToRaf ~ normal(0,10);
        beta_AKTToRaf ~ normal(0,10);
        beta0_RafToMek ~ normal(0,10);
        beta_RafToMek ~ normal(0,10);
        beta0_MekT0Erk ~ normal(0,10);
        beta_MekT0Erk ~ normal(0,10);
        EGF_train ~ normal(mu_EGF, sigma_EGF);
        SOS_train ~ normal(SOS_loc, 5.4);      // likelihood
        Ras_train ~ normal(Ras_loc, 5.3); //ras_loc = 47
        AKT_train ~ normal(AKT_loc,5.1);
        Raf_train ~ normal(Raf_loc,5.3);
        Mek_train ~ normal(Mek_loc,4.7);
        Erk_train ~ normal(Erk_loc,3.2);
    }
"
```

```{r, message=FALSE, warning=FALSE}
mod_misspecify <- rstan::stan_model(model_code = model_str_IGF_misspecify)
```

# Mutilated model

```{r}
num_samples = 500
mutilated_model_misspecify <- function(mu_EGF, sigma_EGF
                                       ,beta0_EGFToSOS, beta_EGFToSOS
                                       ,beta0_SOSToRas, beta_SOSToRas
                                       ,beta0_AKT,beta_EGFToAKT, beta_RasToAKT
                                       ,beta0_RasAKTToRaf, beta_RasToRaf, beta_AKTToRaf
                                       ,beta0_RafToMek, beta_RafToMek
                                       ,beta0_MekT0Erk, beta_MekT0Erk
                                       ,sos
) {
  EGF = rnorm(num_samples, mu_EGF, sigma_EGF)
  SOS = rep(sos, num_samples)
  Ras = 100 / (1 + exp(-beta0_SOSToRas - SOS * beta_SOSToRas)) + rnorm(num_samples,0,5.5)
  AKT = 100 / (1 + exp(-beta0_AKT - EGF * beta_EGFToAKT - Ras * beta_RasToAKT)) + rnorm(n = num_samples,0,5.2)
  Raf = 100 / (1 + exp(-beta0_RasAKTToRaf - Ras * beta_RasToRaf - AKT * beta_AKTToRaf)) + rnorm(n = num_samples,0,5.4)
  Mek = 100 / (1 + exp(-beta0_RafToMek - Raf * beta_RafToMek)) + rnorm(n = num_samples,0,4.9)
  Erk = 100 / (1 + exp(-beta0_MekT0Erk - Mek * beta_MekT0Erk)) + rnorm(n = num_samples,0,3.2)
  return(Erk)
}
```

```{r}
cluster = makeCluster(detectCores() - 1) # convention to leave 1 core for OS
registerDoParallel(cluster)
```

```{r}
#hmc_fit_list_misspecify <- list()
vb_fit_list_misspecify <- list()
ndp = c(140, 160, 200)
start_time <- Sys.time()
for (data_count in 1:K) {
  print(data_count)
  # hmc_fit_list_misspecify[[data_count]] <- foreach (num_data_points=ndp) %dopar% rstan::sampling(mod_misspecify, data=list(D=num_data_points
  #                 ,SOS_train = observational_data_list[[data_count]]$SOS[1:num_data_points]
  #                 ,AKT_train = observational_data_list[[data_count]]$AKT[1:num_data_points]
  #                 ,Raf_train = observational_data_list[[data_count]]$Raf[1:num_data_points]
  #                 ,Mek_train = observational_data_list[[data_count]]$Mek[1:num_data_points]
  #                 ,Erk_train = observational_data_list[[data_count]]$Erk[1:num_data_points]
  #                 ,Ras_train = observational_data_list[[data_count]]$Ras[1:num_data_points]),
  #                 chains = 2, iter = 4000, warmup = 2000, seed = 1, control = list(max_treedepth = 15, adapt_delta = 0.99))
  vb_fit_list_misspecify[[data_count]] <- foreach (num_data_points=ndp) %dopar% rstan::vb(mod_misspecify, data=list(D=num_data_points
                  ,SOS_train = observational_data_list[[data_count]]$SOS[1:num_data_points]
                  ,AKT_train = observational_data_list[[data_count]]$AKT[1:num_data_points]
                  ,Raf_train = observational_data_list[[data_count]]$Raf[1:num_data_points]
                  ,Mek_train = observational_data_list[[data_count]]$Mek[1:num_data_points]
                  ,Erk_train = observational_data_list[[data_count]]$Erk[1:num_data_points]
                  ,Ras_train = observational_data_list[[data_count]]$Ras[1:num_data_points]),
                  algorithm = "fullrank")
}
end_time <- Sys.time()
end_time - start_time
#saveRDS(hmc_fit_list_misspecify,"data/Case3/output/hmc_fit_list_misspecify_Case3.RData") 
```
```{r}
#hmc_fit_list <- list()
vb_fit_list <- list()
start_time <- Sys.time()
for (data_count in 1:K) {
  print(data_count)
  # hmc_fit_list[[data_count]] <- foreach (num_data_points=ndp) %dopar% rstan::sampling(mod, data=list(D=num_data_points
  #                 ,SOS_train = observational_data_list[[data_count]]$SOS[1:num_data_points]
  #                 ,AKT_train = observational_data_list[[data_count]]$AKT[1:num_data_points]
  #                 ,Raf_train = observational_data_list[[data_count]]$Raf[1:num_data_points]
  #                 ,Mek_train = observational_data_list[[data_count]]$Mek[1:num_data_points]
  #                 ,Erk_train = observational_data_list[[data_count]]$Erk[1:num_data_points]
  #                 ,Ras_train = observational_data_list[[data_count]]$Ras[1:num_data_points]),
  #                 chains = 2, iter = 4000, warmup = 2000, seed = 1, control = list(max_treedepth = 15, adapt_delta = 0.99))
  vb_fit_list_misspecify[[data_count]] <- foreach (num_data_points=ndp) %dopar% rstan::vb(mod, data=list(D=num_data_points
                  ,SOS_train = observational_data_list[[data_count]]$SOS[1:num_data_points]
                  ,AKT_train = observational_data_list[[data_count]]$AKT[1:num_data_points]
                  ,Raf_train = observational_data_list[[data_count]]$Raf[1:num_data_points]
                  ,Mek_train = observational_data_list[[data_count]]$Mek[1:num_data_points]
                  ,Erk_train = observational_data_list[[data_count]]$Erk[1:num_data_points]
                  ,Ras_train = observational_data_list[[data_count]]$Ras[1:num_data_points]),
                  algorithm = "fullrank")
}
end_time <- Sys.time()
end_time - start_time
#saveRDS(hmc_fit_list,"data/Case3/output/hmc_fit_list_Case3.RData") 
```

```{r, echo = FALSE}
estimated_y_hmc_list_misspecify <- list()
means_hmc_misspecify <- c()
#means_hmc_list_misspecify <- list()
n=1
for (num_data_points in ndp) {
  for (data_count in 1:K) {
    samples_hmc_misspecify <- rstan::extract(hmc_fit_list_misspecify[[data_count]][[n]], c("mu_EGF", "sigma_EGF"
                                     ,"beta0_EGFToSOS","beta_EGFToSOS"
                                     ,"beta0_SOSToRas","beta_SOSToRas"
                                     ,"beta0_AKT", "beta_EGFToAKT","beta_RasToAKT"
                                     ,"beta0_RasAKTToRaf", "beta_RasToRaf", "beta_AKTToRaf"
                                     ,"beta0_RafToMek", "beta_RafToMek"
                                     ,"beta0_MekT0Erk", "beta_MekT0Erk"
                                     ))
    estimated_y_hmc_list_misspecify[[data_count]] <-  mutilated_model_misspecify(
                                                                          mu_EGF = mean(samples_hmc_misspecify$mu_EGF),
                                                                          sigma_EGF = mean(samples_hmc_misspecify$sigma_EGF),
                                                                          beta0_EGFToSOS = mean(samples_hmc_misspecify$beta0_EGFToSOS),
                                                                          beta_EGFToSOS = mean(samples_hmc_misspecify$beta_EGFToSOS),
                                                                          beta0_SOSToRas = mean(samples_hmc_misspecify$beta0_SOSToRas),
                                                                          beta_SOSToRas = mean(samples_hmc_misspecify$beta_SOSToRas),
                                                                          beta0_AKT = mean(samples_hmc_misspecify$beta0_AKT),
                                                                          beta_EGFToAKT = mean(samples_hmc_misspecify$beta_EGFToAKT),
                                                                          beta_RasToAKT = mean(samples_hmc_misspecify$beta_RasToAKT),
                                                                          beta0_RasAKTToRaf = mean(samples_hmc_misspecify$beta0_RasAKTToRaf),
                                                                          beta_RasToRaf = mean(samples_hmc_misspecify$beta_RasToRaf),
                                                                          beta_AKTToRaf = mean(samples_hmc_misspecify$beta_AKTToRaf),
                                                                          beta0_RafToMek = mean(samples_hmc_misspecify$beta0_RafToMek),
                                                                          beta_RafToMek = mean(samples_hmc_misspecify$beta_RafToMek),
                                                                          beta0_MekT0Erk = mean(samples_hmc_misspecify$beta0_MekT0Erk),
                                                                          beta_MekT0Erk = mean(samples_hmc_misspecify$beta_MekT0Erk),
                                                                          sos = 70
    )
    means_hmc_misspecify <- c(means_hmc_misspecify,abs(mean(estimated_y_hmc_list_misspecify[[data_count]]) - mean(intervention_data_list[[data_count]]$Erk)))

  }
  n = n + 1
  means_hmc_list_misspecify[[paste0("num_data_points_",num_data_points)]] <- means_hmc_misspecify
  means_hmc_misspecify <- c()
}
#saveRDS(means_hmc_list_misspecify,"data/Case3/output/means_hmc_list_misspecify_Case3.RData") 
```


# Plug-in approach

In this section we want to compare our approach with the plug-in estimator. We have to construct a parametric model for the conditional distributions that appear in the formula of an identifying functional of a causal effect. For case study 2, the do-calculus based formula is as follows:

$$
\sum_{Mek, Raf, Ras}P(Erk∣AKT,Mek,Raf,Ras,SOS)P(Mek∣AKT,Raf,Ras,SOS)P(Raf∣AKT,Ras,SOS)P(Ras∣SOS)\sum_{SOS'} P(AKT∣Ras,SOS′)P(SOS′)
$$

In our case, the variables are continuous, hence we have integrals instead of the sum. We can see that we have to make a lot of parametric assumptions about each of the conditional distributions that appear in this formula.

The Stan model is as follows:

```{r}
model_str_IGF_plug_in1 <- "
    data {
        int D;
        vector[D] SOS_train;
        vector[D] AKT_train;
        vector[D] Ras_train;
    }
  parameters {
       real<lower=0> mu_SOS; 
       real<lower=0> sigma_SOS;
       real beta0_AKT;
       real<lower=0> beta_SOSToAKT;
       real<lower=0> beta_RasToAKT;
  }
    transformed parameters {
      vector[D] AKT_loc;
      for (i in 1:D){
        AKT_loc[i] = 100 / (1 + exp(-beta0_AKT - SOS_train[i] * beta_SOSToAKT - Ras_train[i] * beta_RasToAKT));
      }
    }
    model {
        mu_SOS ~ normal(45, 3);
        sigma_SOS ~ normal(5,1);
        beta0_AKT ~ normal(0,10);
        beta_SOSToAKT ~ normal(0,10);
        beta_RasToAKT ~ normal(0,10);
        SOS_train ~ normal(mu_SOS, sigma_SOS);      // likelihood
        AKT_train ~ normal(AKT_loc,5.1);
    }
"
```

```{r}
model_str_IGF_plug_in2 <- "
    data {
        int D;
        vector[D] SOS_train;
        vector[D] AKT_train;
        vector[D] Raf_train;
        vector[D] Mek_train;
        vector[D] Erk_train;
        vector[D] Ras_train;
    }
  parameters {
       real beta0_SOSToRas;
       real<lower=0> beta_SOSToRas;

       real beta0_RasAKTToRaf;
       real<lower=0> beta_RasToRaf;
       real<upper=0> beta_AKTToRaf;
       real<lower=0> beta_SOSToRaf;
       real beta0_RafToMek;
       real<lower=0> beta_RafToMek;
       real<lower=0> beta_AKTToMek;
       real<lower=0> beta_RasToMek;
       real<lower=0> beta_SOSToMek;
       real beta0_MekT0Erk;
       real<lower=0> beta_MekT0Erk;
       real<lower=0> beta_AKTT0Erk;
       real<lower=0> beta_RafT0Erk;
       real<lower=0> beta_RasT0Erk;
       real<lower=0> beta_SOST0Erk;
  }
    transformed parameters {
      vector[D] Ras_loc;
      vector[D] Raf_loc;
      vector[D] Mek_loc;
      vector[D] Erk_loc;
      for (i in 1:D){
        Ras_loc[i] = 100 / (1 + exp(-beta0_SOSToRas -SOS_train[i] * beta_SOSToRas));
        Raf_loc[i] = 100 / (1 + exp(-beta0_RasAKTToRaf - Ras_train[i] * beta_RasToRaf - AKT_train[i] * beta_AKTToRaf - SOS_train[i] * beta_SOSToRaf));
        Mek_loc[i] = 100 / (1 + exp(-beta0_RafToMek - Raf_train[i] * beta_RafToMek - AKT_train[i] * beta_AKTToMek - Ras_train[i] * beta_RasToMek - SOS_train[i] * beta_SOSToMek));
        Erk_loc[i] = 100 / (1 + exp(-beta0_MekT0Erk - Mek_train[i] * beta_MekT0Erk - AKT_train[i] * beta_AKTT0Erk - Raf_train[i] * beta_RafT0Erk - Ras_train[i] * beta_RasT0Erk - SOS_train[i] * beta_SOST0Erk));
      }
    }
    model {
        beta0_SOSToRas ~ normal(0,10);
        beta_SOSToRas ~ normal(0,10);
        beta0_RasAKTToRaf ~ normal(0,10);
        beta_RasToRaf ~ normal(0,10);
        beta_AKTToRaf ~ normal(0,10);
        beta_SOSToRaf ~ normal(0,10);
        beta0_RafToMek ~ normal(0,10);
        beta_RafToMek ~ normal(0,10);
        beta_AKTToMek ~ normal(0,10);
        beta_RasToMek ~ normal(0,10);
        beta_SOSToMek ~ normal(0,10);
        beta0_MekT0Erk ~ normal(0,10);
        beta_MekT0Erk ~ normal(0,10);
        beta_AKTT0Erk ~ normal(0,10);
        beta_RafT0Erk ~ normal(0,10);
        beta_RasT0Erk ~ normal(0,10);
        beta_SOST0Erk ~ normal(0,10);
        Ras_train ~ normal(Ras_loc, 5.3); //ras_loc = 47
        Raf_train ~ normal(Raf_loc,5.3);
        Mek_train ~ normal(Mek_loc,4.7);
        Erk_train ~ normal(Erk_loc,3.2);
    }
"
```


```{r, message=FALSE, warning=FALSE}
mod_plug_in1 <- rstan::stan_model(model_code = model_str_IGF_plug_in1)
```

```{r, message=FALSE, warning=FALSE}
mod_plug_in2 <- rstan::stan_model(model_code = model_str_IGF_plug_in2)
```

# Mutilated model

```{r}
num_samples = 500
mutilated_model_plug_in1 <- function(mu_SOS, sigma_SOS,
                                     beta0_AKT, beta_SOSToAKT, beta_RasToAKT,
                                     data_count) {
  SOS = rnorm(num_samples, mu_SOS, sigma_SOS)
  Ras = observational_data_list[[data_count]]$Ras[1:num_samples]
  AKT = 100 / (1 + exp(-beta0_AKT - SOS * beta_SOSToAKT - Ras * beta_RasToAKT)) + rnorm(n = num_samples,0,5.2)
  return(AKT)
}
```

```{r}
num_samples = 500
mutilated_model_plug_in <- function(mu_SOS, sigma_SOS,
                                    beta0_SOSToRas, beta_SOSToRas,
                                    beta0_AKT, beta_SOSToAKT, beta_RasToAKT,
                                    beta0_RasAKTToRaf, beta_RasToRaf, beta_AKTToRaf, beta_SOSToRaf,
                                    beta0_RafToMek, beta_RafToMek, beta_AKTToMek, beta_RasToMek, beta_SOSToMek,
                                    beta0_MekT0Erk, beta_MekT0Erk, beta_AKTT0Erk, beta_RafT0Erk, beta_RasT0Erk, beta_SOST0Erk
                                    ,sos) {
  SOS = rep(sos, num_samples)
  Ras = 100 / (1 + exp(-beta0_SOSToRas - SOS * beta_SOSToRas)) + rnorm(num_samples,0,5.5)
  AKT = mutilated_model_plug_in1(mu_SOS, sigma_SOS,
                                     beta0_AKT, beta_SOSToAKT, beta_RasToAKT,
                                     data_count)
  Raf = 100 / (1 + exp(-beta0_RasAKTToRaf - Ras * beta_RasToRaf - AKT * beta_AKTToRaf- SOS * beta_SOSToRaf)) + rnorm(n = num_samples,0,5.4)
  Mek = 100 / (1 + exp(-beta0_RafToMek - Raf * beta_RafToMek- AKT * beta_AKTToMek - Ras * beta_RasToMek - SOS * beta_SOSToMek)) + rnorm(n = num_samples,0,4.9)
  Erk = 100 / (1 + exp(-beta0_MekT0Erk - Mek * beta_MekT0Erk - AKT * beta_AKTT0Erk - Raf * beta_RafT0Erk - Ras * beta_RasT0Erk - SOS * beta_SOST0Erk)) + rnorm(n = num_samples,0,3.2)
  return(Erk)
}
```

```{r, echo = FALSE}
estimated_y_hmc_list_plug_in <- list()
hmc_fit_list_plug_in <- list()
means_hmc_plug_in <- c()
means_hmc_list_plug_in <- list()
start_time <- Sys.time()
for (num_data_points in ndp) {
  for (data_count in 1:K) {
    data_list <- list(D=num_data_points
                  ,SOS_train = observational_data_list[[data_count]]$SOS[1:num_data_points]
                  ,AKT_train = observational_data_list[[data_count]]$AKT[1:num_data_points]
                  ,Raf_train = observational_data_list[[data_count]]$Raf[1:num_data_points]
                  ,Mek_train = observational_data_list[[data_count]]$Mek[1:num_data_points]
                  ,Erk_train = observational_data_list[[data_count]]$Erk[1:num_data_points]
                  ,Ras_train = observational_data_list[[data_count]]$Ras[1:num_data_points]
                  )
    
    #If you get this error (Cluster setup failed. 2 of 2 workers failed to connect.) run these lines:
    if (Sys.getenv("RSTUDIO") == "1" && !nzchar(Sys.getenv("RSTUDIO_TERM")) &&
        Sys.info()["sysname"] == "Darwin") {
      parallel:::setDefaultClusterOptions(setup_strategy = "sequential")
    }
    hmc_fit_list_plug_in[[data_count]] <- rstan::sampling(mod_plug_in, data=data_list, chains = 2, iter = 4000, warmup = 2000, seed = 1, control = list(max_treedepth = 15, adapt_delta = 0.99))
    samples_hmc_plug_in <- rstan::extract(hmc_fit_list_plug_in[[data_count]], c("mu_SOS", "sigma_SOS"
                                     ,"beta0_SOSToRas","beta_SOSToRas"
                                     ,"beta0_AKT", "beta_SOSToAKT", "beta_RasToAKT"
                                     ,"beta0_RasAKTToRaf", "beta_RasToRaf", "beta_AKTToRaf","beta_SOSToRaf"
                                     ,"beta0_RafToMek", "beta_RafToMek","beta_AKTToMek","beta_RasToMek","beta_SOSToMek"
                                     ,"beta0_MekT0Erk", "beta_MekT0Erk","beta_AKTT0Erk","beta_RafT0Erk","beta_RasT0Erk","beta_SOST0Erk"
                                     ))
    estimated_y_hmc_list_plug_in[[data_count]] <-  mutilated_model_plug_in(
                                                                          beta0_SOSToRas = mean(samples_hmc_plug_in$beta0_SOSToRas),
                                                                          beta_SOSToRas = mean(samples_hmc_plug_in$beta_SOSToRas),
                                                                          beta0_AKT = mean(samples_hmc_plug_in$beta0_AKT),
                                                                          beta_SOSToAKT = mean(samples_hmc_plug_in$beta_SOSToAKT),
                                                                          beta_RasToAKT = mean(samples_hmc_plug_in$beta_RasToAKT),
                                                                          beta0_RasAKTToRaf = mean(samples_hmc_plug_in$beta0_RasAKTToRaf),
                                                                          beta_RasToRaf = mean(samples_hmc_plug_in$beta_RasToRaf),
                                                                          beta_AKTToRaf = mean(samples_hmc_plug_in$beta_AKTToRaf),
                                                                          beta_SOSToRaf = mean(samples_hmc_plug_in$beta_SOSToRaf),
                                                                          beta0_RafToMek = mean(samples_hmc_plug_in$beta0_RafToMek),
                                                                          beta_RafToMek = mean(samples_hmc_plug_in$beta_RafToMek),
                                                                          beta_AKTToMek = mean(samples_hmc_plug_in$beta_AKTToMek),
                                                                          beta_RasToMek = mean(samples_hmc_plug_in$beta_RasToMek),
                                                                          beta_SOSToMek = mean(samples_hmc_plug_in$beta_SOSToMek),
                                                                          beta0_MekT0Erk = mean(samples_hmc_plug_in$beta0_MekT0Erk),
                                                                          beta_MekT0Erk = mean(samples_hmc_plug_in$beta_MekT0Erk),
                                                                          beta_AKTT0Erk = mean(samples_hmc_plug_in$beta_AKTT0Erk),
                                                                          beta_RafT0Erk = mean(samples_hmc_plug_in$beta_RafT0Erk),
                                                                          beta_RasT0Erk = mean(samples_hmc_plug_in$beta_RasT0Erk),
                                                                          beta_SOST0Erk = mean(samples_hmc_plug_in$beta_SOST0Erk),
                                                                          sos = 70
    )
    means_hmc_plug_in <- c(means_hmc_plug_in,abs(mean(estimated_y_hmc_list_plug_in[[data_count]]) - mean(intervention_data_list[[data_count]]$Erk)))
   }
  means_hmc_list_plug_in[[paste0("num_data_points_",num_data_points)]] <- means_hmc_plug_in
  means_hmc_plug_in <- c()
}
end_time <- Sys.time()
end_time - start_time
```
