---
title: "Case Study 2 : The napkin problem"
author: "Sara Taheri"
date: "11/24/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(rstan)
library(ggplot2)
library(dplyr)
```

```{r}
D = 300
startSeed = 250
x = rep(1,D)
K = 10
ndp = c(15,20, 30, 60, 100)
options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)
```

```{r}
cluster = makeCluster(detectCores() - 1) # convention to leave 1 core for OS
registerDoParallel(cluster)
```

# Generate observational data

```{r}
#True parameters
mu_UXW = 3
sigma_UXW = 1
mu_UWY = 5
sigma_UWY = 1

beta_UXW_to_W = 0.3
beta_UWY_to_W = 0.5

beta_W_to_R = 0.7

beta_zero_X = 2
beta_UXW_to_X = 0.23
beta_R_to_X = 0.1

beta_UWY_to_Y = 0.2
beta_X_to_Y = 0.4

```


```{r}
# Create observational data
obs_data_list <- list()
data_count = 1
for (seed in startSeed:(startSeed+K-1)) {
  set.seed(seed)
  UXW = rnorm(D, mean = mu_UXW, sd = sigma_UXW)
  UWY = rnorm(D, mean = mu_UWY, sd = sigma_UWY)
  W =  rgamma(D, shape = 1/(5*(UXW * beta_UXW_to_W + beta_UWY_to_W * UWY)^2), scale = 5*(UXW * beta_UXW_to_W + beta_UWY_to_W * UWY))
  R =  rnorm(D, W * beta_W_to_R, 6)
  
  p1_train = rep(0,D)
  p1_train = 1/(1+exp(-beta_zero_X - beta_UXW_to_X * UXW - beta_R_to_X * R))
  X <- rbinom(n = D, size = 1, prob = p1_train)
  
  Y =  rnorm(D, UWY * beta_UWY_to_Y + X * beta_X_to_Y, 6)
  d = data.frame("W" = W, "R" = R, "X" = X, "Y" = Y)
  obs_data_list[[data_count]] = d
  data_count = data_count + 1
}
```

# Generate interventional data

```{r}
# Create interventional data
intv_data_list <- list()
data_count = 1
for (seed in startSeed:(startSeed+K-1)) {
  set.seed(seed)
  X_intv = x
  Y_intv = rnorm(D,UWY * beta_UWY_to_Y + X_intv * beta_X_to_Y,6)
  intv_data_list[[data_count]] = data.frame("W" = obs_data_list[[data_count]]$W, "R" = obs_data_list[[data_count]]$R, "X" = X_intv, "Y" = Y_intv)
  data_count = data_count + 1
}
```


# Napkin case study Stan model

```{r}
model_str_napkin <- "
    data {
        int D;
        vector[D] W;
        vector[D] R;
        int<lower=0,upper=1> X[D];
        vector[D] Y;
    }
  parameters {
       real mu_UXW; 
       real<lower=0> sigma_UXW;
       real mu_UWY; 
       real<lower=0> sigma_UWY;
       real beta_UXW_to_W; 
       real beta_UWY_to_W;
       real beta_W_to_R;
       real beta_zero_X;
       real beta_UXW_to_X;
       real beta_R_to_X;
       real beta_UWY_to_Y;
       real beta_X_to_Y;
       vector[D] UXW;
       vector[D] UWY;
    }
    transformed parameters {
      vector[D] W_shape;
      vector[D] W_scale;
      vector[D] R_loc;
      vector[D] Y_loc;
      for (i in 1:D){
        W_shape[i] = 1/(5*(UXW[i] * beta_UXW_to_W + beta_UWY_to_W * UWY[i])*(UXW[i] * beta_UXW_to_W + beta_UWY_to_W * UWY[i])); 
        W_scale[i] = fabs(5*(UXW[i] * beta_UXW_to_W + beta_UWY_to_W * UWY[i]));
        R_loc[i] = W[i] * beta_W_to_R;
        Y_loc[i] = UWY[i] * beta_UWY_to_Y + X[i] * beta_X_to_Y;
      }
    }
    model {
        mu_UXW  ~ normal(0, 10); 
        sigma_UXW  ~ normal(0, 10);
        mu_UWY  ~ normal(0, 10);
        sigma_UWY ~ normal(0, 10);
        beta_UXW_to_W ~ normal(0,10);
        beta_UWY_to_W ~ normal(0, 10);
        beta_W_to_R ~ normal(0,10);
        beta_zero_X ~ normal(0,10);
        beta_UXW_to_X ~ normal(0,10);
        beta_R_to_X ~ normal(0,10);
        beta_UWY_to_Y ~ normal(0,10);
        beta_X_to_Y ~ normal(0,10);
        W ~ gamma(W_shape, W_scale);
        R ~ normal(R_loc, 6);
        X ~ bernoulli_logit(beta_zero_X + beta_UXW_to_X * UXW + beta_R_to_X * R);
        Y ~ normal(Y_loc, 6); 
        UXW ~ normal(mu_UXW,sigma_UXW);
        UWY ~ normal(mu_UWY,sigma_UWY);
    }
"
```

```{r, message=FALSE, warning=FALSE}
mod <- rstan::stan_model(model_code = model_str_napkin)
```

# Mutilated model

```{r}
mutilated_model <- function(mu_UWY, sigma_UWY, beta_UWY_to_Y, beta_X_to_Y, x, num_data_points, seed) {
  set.seed(seed)
  UWY = rnorm(num_data_points, mu_UWY, sigma_UWY)
  X = x[1:num_data_points]
  Y = rnorm(num_data_points,
            UWY * beta_UWY_to_Y + X * beta_X_to_Y
            , 6)
  return(Y)
}
```


```{r}
hmc_fit_list <- list()
start_time <- Sys.time()
for (data_count in 1:K) {
    print(data_count)
    hmc_fit_list[[data_count]] <- foreach (num_data_points=ndp) %dopar% rstan::sampling(mod, 
                                                                                        data = list(D=num_data_points,
                                                                                                    W=obs_data_list[[data_count]]$W[1:num_data_points],
                                                                                                    R=obs_data_list[[data_count]]$R[1:num_data_points],
                                                                                                    X=obs_data_list[[data_count]]$X[1:num_data_points],
                                                                                                    Y=obs_data_list[[data_count]]$Y[1:num_data_points]),
                                                                                        chains = 2,
                                                                                        iter = 3000,
                                                                                        warmup = 1500,
                                                                                        seed = 1,
                                                                                        control = list(max_treedepth = 15))
    
}
end_time <- Sys.time()
end_time - start_time
#saveRDS(hmc_fit_list,"data/Case2/output/hmc_fit_list_Case2.RData") 
```

```{r}
estimated_y_hmc_list <- list()
means_hmc <- c()
means_hmc_list <- list()
n=1

for (num_data_points in ndp) {
  for (data_count in 1:K) {
    samples_hmc <- rstan::extract(hmc_fit_list[[data_count]][[n]], c("mu_UXW","sigma_UXW", "mu_UWY", "sigma_UWY", "beta_UXW_to_W", "beta_UWY_to_W", "beta_W_to_R",
                                                                     "beta_zero_X", "beta_UXW_to_X", "beta_R_to_X",
                                                                     "beta_UWY_to_Y", "beta_X_to_Y"))
    estimated_y_hmc_list[[data_count]] <-  mutilated_model(mu_UWY = mean(samples_hmc$mu_UWY),
                                                           sigma_UWY = mean(samples_hmc$sigma_UWY),
                                                           beta_UWY_to_Y = mean(samples_hmc$beta_UWY_to_Y),
                                                           beta_X_to_Y = mean(samples_hmc$beta_X_to_Y),
                                                           x = x,
                                                           num_data_points = num_data_points,
                                                           seed = startSeed
    )
    means_hmc <- c(means_hmc,abs(mean(estimated_y_hmc_list[[data_count]]) - mean(intv_data_list[[data_count]]$Y)))
  }
  n=n+1
  means_hmc_list[[paste0("num_data_points_",num_data_points)]] <- means_hmc
  means_hmc <- c()
}
#saveRDS(means_hmc_list, file = "data/Case2/output/means_hmc_list_Case2.RData")
```

# Check model misspecification

Let's assume that we misspecified the model by assuming that there are 2 unobserved confounders between W and Y.

# Napkin case study misspecified Stan model

```{r}
model_str_napkin_misspecify <- "
    data {
        int D;
        vector[D] W;
        vector[D] R;
        int<lower=0,upper=1> X[D];
        vector[D] Y;
    }
  parameters {
       real mu_UXW; 
       real<lower=0> sigma_UXW;
       real mu_UWY1; 
       real<lower=0> sigma_UWY1;
       real mu_UWY2; 
       real<lower=0> sigma_UWY2;
       real beta_UXW_to_W; 
       real beta_UWY1_to_W;
       real beta_UWY2_to_W;
       real beta_W_to_R;
       real beta_zero_X;
       real beta_UXW_to_X;
       real beta_R_to_X;
       real beta_UWY1_to_Y;
       real beta_UWY2_to_Y;
       real beta_X_to_Y;
       vector[D] UXW;
       vector[D] UWY1;
       vector[D] UWY2;
    }
    transformed parameters {
      vector[D] W_shape;
      vector[D] W_scale;
      vector[D] R_loc;
      vector[D] Y_loc;
      for (i in 1:D){
        W_shape[i] = 1/(5*(UXW[i] * beta_UXW_to_W + beta_UWY1_to_W * UWY1[i] + beta_UWY2_to_W * UWY2[i])*(UXW[i] * beta_UXW_to_W + beta_UWY1_to_W * UWY1[i] + beta_UWY2_to_W * UWY2[i])); 
        W_scale[i] = fabs(5*(UXW[i] * beta_UXW_to_W + beta_UWY1_to_W * UWY1[i] + beta_UWY2_to_W * UWY2[i]));
        R_loc[i] = W[i] * beta_W_to_R;
        Y_loc[i] = UWY1[i] * beta_UWY1_to_Y + UWY2[i] * beta_UWY2_to_Y + X[i] * beta_X_to_Y;
      }
    }
    model {
        mu_UXW  ~ normal(0, 10); 
        sigma_UXW  ~ normal(0, 10);
        mu_UWY1  ~ normal(0, 10);
        sigma_UWY1 ~ normal(0, 10);
        mu_UWY2  ~ normal(0, 10);
        sigma_UWY2 ~ normal(0, 10);
        beta_UXW_to_W ~ normal(0,10);
        beta_UWY1_to_W ~ normal(0, 10);
        beta_UWY2_to_W ~ normal(0, 10);
        beta_W_to_R ~ normal(0,10);
        beta_zero_X ~ normal(0,10);
        beta_UXW_to_X ~ normal(0,10);
        beta_R_to_X ~ normal(0,10);
        beta_UWY1_to_Y ~ normal(0,10);
        beta_UWY2_to_Y ~ normal(0,10);
        beta_X_to_Y ~ normal(0,10);
        W ~ gamma(W_shape, W_scale);
        R ~ normal(R_loc, 6);
        X ~ bernoulli_logit(beta_zero_X + beta_UXW_to_X * UXW + beta_R_to_X * R);
        Y ~ normal(Y_loc, 6); 
        UXW ~ normal(mu_UXW,sigma_UXW);
        UWY1 ~ normal(mu_UWY1,sigma_UWY1);
        UWY2 ~ normal(mu_UWY2,sigma_UWY2);
    }
"
```

```{r, message=FALSE, warning=FALSE}
mod_misspecify <- rstan::stan_model(model_code = model_str_napkin_misspecify)
```

# Mutilated model

```{r}
mutilated_model_misspecify <- function(mu_UWY1, sigma_UWY1, mu_UWY2, sigma_UWY2, beta_UWY1_to_Y, beta_UWY2_to_Y, beta_X_to_Y, x, num_data_points, seed) {
  set.seed(seed)
  UWY1 = rnorm(num_data_points, mu_UWY1, sigma_UWY1)
  UWY2 = rnorm(num_data_points, mu_UWY2, sigma_UWY2)
  X = x[1:num_data_points]
  Y = rnorm(num_data_points,
            UWY1 * beta_UWY1_to_Y +  UWY2 * beta_UWY2_to_Y + X * beta_X_to_Y
            , 6)
  return(Y)
}
```

```{r}
hmc_fit_list_misspecify <- list()
start_time <- Sys.time()
for (data_count in 1:K) {
  print(data_count)
  hmc_fit_list_misspecify[[data_count]] <- foreach (num_data_points=ndp) %dopar% rstan::sampling(mod_misspecify,
                                                                                                 data = list(D=num_data_points,
                                                                                                             W=obs_data_list[[data_count]]$W[1:num_data_points],
                                                                                                             R=obs_data_list[[data_count]]$R[1:num_data_points],
                                                                                                             X=obs_data_list[[data_count]]$X[1:num_data_points],
                                                                                                             Y=obs_data_list[[data_count]]$Y[1:num_data_points]),
                                                                                                 chains = 2,
                                                                                                 iter = 3000,
                                                                                                 warmup = 1500,
                                                                                                 seed = 1,
                                                                                                 control = list(max_treedepth = 15))

}
end_time <- Sys.time()
end_time - start_time
#saveRDS(hmc_fit_list_misspecify,"data/Case2/output/hmc_fit_list_misspecify_Case2.RData") 
```

```{r}
estimated_y_hmc_list_misspecify <- list()
means_hmc_misspecify <- c()
means_hmc_list_misspecify <- list()
n=1

for (num_data_points in ndp) {
  for (data_count in 1:K) {
    samples_hmc_misspecify <- rstan::extract(hmc_fit_list_misspecify[[data_count]][[n]],
    c("mu_UWY1", "sigma_UWY1", "mu_UWY2", "sigma_UWY2", "beta_UWY1_to_Y", "beta_UWY2_to_Y", "beta_X_to_Y"))
    estimated_y_hmc_list_misspecify[[data_count]] <-  mutilated_model_misspecify(mu_UWY = mean(samples_hmc_misspecify$mu_UWY),
                                                           sigma_UWY = mean(samples_hmc_misspecify$sigma_UWY),
                                                           beta_UWY_to_Y = mean(samples_hmc_misspecify$beta_UWY_to_Y),
                                                           beta_X_to_Y = mean(samples_hmc_misspecify$beta_X_to_Y),
                                                           x = x,
                                                           num_data_points = num_data_points,
                                                           seed = startSeed
    )
    means_hmc_misspecify <- c(means_hmc_misspecify,abs(mean(estimated_y_hmc_list_misspecify[[data_count]]) - mean(intv_data_list[[data_count]]$Y)))
  }
  n=n+1
  means_hmc_list_misspecify[[paste0("num_data_points_",num_data_points)]] <- means_hmc_misspecify
  means_hmc_misspecify <- c()
}
#saveRDS(means_hmc_list_misspecify, file = "data/Case2/output/means_hmc_list_misspecify_Case2.RData")
```


# Model Plug in

In this section we want to compare our approach with the plug-in estimator. We have to construct a parametric model for the conditional distributions that appear in the formula of an identifying functional of a causal effect. For case study 4, the do-calculus based formula is as follows:

$$
P(Y|do(x)) = \frac{\int_{W}P(X,Y∣R,W)P(W)dW}{\int_{W}P(X|R,W)P(W)dW} = \frac{\int_{W}P(Y|X,W,R)P(X|R,W)P(W)dW}{\int_{W}P(X|R,W)P(W)dW}
$$

```{r}
model_str_napkin_plug_in <- "
    data {
        int D;
        vector[D] W;
        vector[D] R;
        int<lower=0,upper=1> X[D];
        vector[D] Y;
    }
  parameters {
       real<lower=0> W_shape;
       real<lower=0> W_scale; 
       real beta_zero_X;
       real beta_R_to_X;
       real beta_W_to_X;
       real beta_W_to_Y;
       real beta_R_to_Y;
       real beta_X_to_Y;
    }
    transformed parameters {
      vector[D] Y_loc;
      for (i in 1:D){
        Y_loc[i] = W[i] * beta_W_to_Y + R[i] * beta_R_to_Y + X[i] * beta_X_to_Y;
      }
    }
    model {
        beta_zero_X ~ normal(0,10);
        beta_R_to_X ~ normal(0, 10);
        beta_W_to_X ~ normal(0, 10);
        beta_W_to_Y ~ normal(0,10);
        beta_R_to_Y ~ normal(0,10);
        beta_X_to_Y ~ normal(0,10);
        W ~ gamma(W_shape, W_scale);
        //R ~ normal(R_loc, 6);
        X ~ bernoulli_logit(beta_zero_X + beta_R_to_X * R + beta_W_to_X * W);
        Y ~ normal(Y_loc, 6); 
    }
"
```

```{r, message=FALSE, warning=FALSE}
mod_plug_in <- rstan::stan_model(model_code = model_str_napkin_plug_in)
```

# Mutilated model

```{r}
mutilated_model_plug_in <- function(W_shape, W_scale, beta_W_to_Y, beta_R_to_Y, beta_X_to_Y, beta_zero_X, beta_R_to_X, beta_W_to_X, x, num_data_points, seed, data_count, obs_data_list) {
  set.seed(seed)
  denominator = 0
  numerator = 0
  p1 = rep(0,num_data_points)
  
  for (i in 1:num_data_points) {
    p1 = 1/(1+exp(-beta_zero_X - beta_R_to_X * obs_data_list[[data_count]]$R[i] - beta_W_to_X * obs_data_list[[data_count]]$W[i]))
    denominator = denominator + (dgamma(x=obs_data_list[[data_count]]$W[i], shape = W_shape, scale = W_scale) * dbinom(x = x, size = 1, prob = p1))
    numerator = numerator + (dgamma(x=obs_data_list[[data_count]]$W[i], shape = W_shape, scale = W_scale) * dbinom(x = x, size = 1, prob = p1) * dnorm(x = obs_data_list[[data_count]]$Y[i], mean = obs_data_list[[data_count]]$W[i] * beta_W_to_Y +  obs_data_list[[data_count]]$R[i] * beta_R_to_Y + x * beta_X_to_Y, sd = 6))
  }
  return(numerator/denominator)
}
```

```{r}
hmc_fit_list_plug_in <- list()
start_time <- Sys.time()
for (data_count in 1:K) {
  print(data_count)
  hmc_fit_list_plug_in[[data_count]] <- foreach (num_data_points=ndp) %dopar% rstan::sampling(mod_plug_in,
                                                                                              data = list(D=num_data_points,
                                                                                                          W=obs_data_list[[data_count]]$W[1:num_data_points],
                                                                                                          R=obs_data_list[[data_count]]$R[1:num_data_points],
                                                                                                          X=obs_data_list[[data_count]]$X[1:num_data_points],
                                                                                                          Y=obs_data_list[[data_count]]$Y[1:num_data_points]),
                                                                                              chains = 2,
                                                                                              iter = 3000,
                                                                                              warmup = 1500,
                                                                                              seed = 1,
                                                                                              control = list(max_treedepth = 15))
}
end_time <- Sys.time()
end_time - start_time
#saveRDS(hmc_fit_list_plug_in,"data/Case2/output/hmc_fit_list_plug_in_Case2.RData") 
```

```{r}
estimated_y_hmc_list_plug_in <- list()
means_hmc_plug_in <- c()
means_hmc_list_plug_in <- list()
n=1

for (num_data_points in ndp) {
  for (data_count in 1:K) {
      samples_hmc_plug_in <- rstan::extract(hmc_fit_list_plug_in[[data_count]][[n]],
                                            c("W_shape", "W_scale", "beta_zero_X", "beta_R_to_X", "beta_W_to_X", "beta_W_to_Y", "beta_R_to_Y", "beta_X_to_Y"))
      estimated_y_hmc_list_plug_in[[data_count]] <-  mutilated_model_plug_in(W_shape = mean(samples_vb_plug_in$W_shape),
                                                                             W_scale = mean(samples_vb_plug_in$W_scale),
                                                                             beta_zero_X = mean(samples_vb_plug_in$beta_zero_X),
                                                                             beta_R_to_X = mean(samples_vb_plug_in$beta_R_to_X),
                                                                             beta_W_to_X = mean(samples_vb_plug_in$beta_W_to_X),
                                                                             beta_W_to_Y = mean(samples_vb_plug_in$beta_W_to_Y),
                                                                             beta_R_to_Y = mean(samples_vb_plug_in$beta_R_to_Y),
                                                                             beta_X_to_Y = mean(samples_vb_plug_in$beta_X_to_Y),
                                                                             x = x,
                                                                             num_data_points = num_data_points,
                                                                             seed = startSeed,
                                                                             data_count = data_count,
                                                                             obs_data_list = obs_data_list
      )
      means_hmc_plug_in <- c(means_hmc_plug_in,abs(mean(estimated_y_hmc_list_plug_in[[data_count]]) - mean(intv_data_list[[data_count]]$Y)))
  }
  n=n+1
  means_hmc_list_plug_in[[paste0("num_data_points_",num_data_points)]] <- means_hmc_plug_in
  means_hmc_plug_in <- c()
}
#saveRDS(means_hmc_list_plug_in, file = "data/Case2/output/means_hmc_list_plug_in_Case2.RData")
```
